Multiple Dice Working as One: CAD Flows and Routing Architectures for Silicon Interposer FPGAs

### 2014-12-29 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
Updating table caption
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4552 8e3573b8-cf2c-4f14-ef6d-137439e28b8b
Fix typos
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4566 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\author{
		% 1st author
		Ehsan~Nasiri,
		% 2nd author
		Javeed~Shaikh,
		% 3rd author
		Andre~Hahn~Pereira,
		% 4th author
        and~Vaughn~Betz,~\IEEEmembership{Member,~IEEE,}%
\thanks{E. Nasiri is with Altera Corporation, Toronto Technology Center, Toronto, Ontario, Canada (email: enasiri@altera.com)}% 
\thanks{J. Shaikh is with Google, Mountain View, California, USA (email: jvds@google.com)}%
\thanks{A. H. Pereira is with the Computer and Digital Systems Engineering Department - Escola Polit\'{e}cnica, University of S\~{a}o Paulo, S\~{a}o Paulo, SP, Brazil (email: andre.hahn@usp.br)}%
\thanks{V. Betz is with the Department of Electrical and Computer Engineering, University of Toronto, Toronto, Ontario, Canada (email: vaughn@eecg.utoronto.ca)}%
\thanks{Manuscript received December 19, 2014.}}

@@

% As a general rule, do not put math, special symbols or citations in the abstract or keywords.
\begin{abstract}
Multiple FPGA dice can be connected through a silicon interposer to form a larger FPGA system than one monolithic die can accomodate. Such systems are now commercially available but many open questions remain concerning their architecture and efficiency, as the number of signals passing between dice is reduced and the signal delay between dice is increased compared to a monolithic FPGA.
We create a new version of VPR to target such systems and investigate modifications to placement and routing and the incorporation of partitioning into the flow to best exploit interposer-based FPGAs. Our best CAD flow can reduce the routing demand for interposer FPGAs with realistic amounts of connectivity between dice by 47\% and improve the circuit speed by 13\% on average. Modifications to the FPGA architecture to add routing flexibility when crossing the interposer are very beneficial and can improve routability by a further 11\%. With these CAD and architecture enhancements, we find that if an interposer supplies (between dice) 40\% of the routing capacity that the normal (within-die) FPGA routing channels supply there is negligible impact on the routability of circuits. Smaller interposer routing capacities do impact routability however: minimum channel width increases by 10\% and 70\% when an interposer supplies only 20\% and 10\% of the within-die routing, respectively. The interposer also impacts delay, increasing circuit delay by 11\% on average for a 1 ns interposer signal delay and a two-die system. Reducing the interposer delay has a greater benefit in improving circuit speed than does reducing the number of dice in the system.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Algorithms, Design, FPGA, Silicon interposer, 2.5D ICs. 
\end{IEEEkeywords}

\section{Introduction}
\label{introSection} 
%   - Describe multi-chip
Interposer-based FPGA systems are composed of multiple FPGA dice connected through a silicon interposer. The interposer is fabricated with an older technology than that used for the dice, and links between dice include both a micro-bump on each FPGA die and a metal wire on the interposer~\cite{xilinxTSVperformance}. This results in a reduced connectivity between dice and increased delay for connections between dice compared to the total routing connectivity and the delay one can achieve within a single die. In this work we present an architecture study of silicon interposer-based FPGAs that analyzes how the reduced connectivity and increased delay impact their performance.

%   - Motivate: yield, bigger FPGA, no previous study
Silicon interposer FPGA technology is interesting because it enables the creation of large FPGAs composed of small dice and also of very large FPGAs, with higher logic capacity than one can achieve with a single die. Such interposer systems are sometimes called 2.5D devices, since they make use of vertical stacking of dice on an interposer to enable higher integration levels~\cite{3dhandbook}.

% yield rate
Making large FPGAs from multiple smaller dice is particularly interesting at the beginning of a new manufacturing process, when defect densities are high. In such a case, good-die yield drops dramatically as the die size increases, and this drastically impacts the availability of large FPGAs early in the process lifetime.

@@

%   - Contributions
The main contributions of our work are as follows:
\begin{itemize}
	\item Modifications of the VPR CAD tool to model and target 2.5D silicon interposer-based FPGAs, as well as CAD changes to optimize for 2.5D FPGAs.
	\item An investigation of the efficacy of incorporating partitioning into various stages of the CAD flow.
	\item A detailed study of how the multi-FPGA system routability is impacted by the amount of between-die connectivity provided by the interposer, and by the switching flexibility at the FPGA die - interposer boundary.
	\item An analysis of the interposer's impact on timing.
\end{itemize}

The paper is organized as follows. Section \ref{siliconSection} gives more information about silicon interposer technology. Section \ref{archSection} describes the changes made to the FPGA architecture description and to VPR to target 2.5D FPGAs. Section \ref{cadSection} describes optimizations made to the VPR placement and routing algorithms to improve quality on such FPGAs and Section~\ref{sec:CADflow} describes how partitioning can be used to further enhance the CAD flow. Section \ref{resultsSection} presents architectural results for 2.5D FPGAs.

An earlier and less detailed version of this work appeared in \cite{interposer2014}. We extend this earier work by enhancing the CAD tools, incorporating partitioning into the flows and investigating a broader range of architecture questions.

@@

Interposer-based FPGAs allow the creation of chips larger than a single die, making a "More than Moore" improvement on the size and number of logic cells possible, and with chips combined far more tightly and with more connectivity than if they were on separate boards connected through conventional means.

The improvement in the capacity of 2.5D FPGAs over conventional ones is very significant. Xilinx's largest 28 nm interposer-based FPGA, the Virtex-7 XC7V2000T, contains 4 FPGA dice (which Xilinx calls Super Logic Regions) and 1.954 million logic cells~\cite{xilinx7series}, while the largest non-interposer Virtex-7 die (the XC7VX980T), has 979k logic cells and Altera's largest 28 nm FPGA, the Stratix V 5SGXBB, has 952k logic elements~\cite{stratixV}. Even though all these FPGAs use a 28 nm process, silicon interposer technology allows the creation of FPGAs with twice the resources possible on even an extremely large single die. Xilinx's recently announced Virtex UltraScale line of 20 nm FPGAs makes even more aggressive use of silicon interposers; the largest monolithic FPGA in this family contains 941k logic cells, while the four largest capacity FPGAs are all built with silicon interposers and span a range from 1.25 to 4.43 million logic cells~\cite{virtexultrascale}.

Another major advantage of interposer-based FPGAs comes at the beginning of a new manufacturing process, when the defect density is high~\cite{xilinxTSVperformance}. Bigger dice suffer a much reduced yield compared to smaller dice, and this greatly affects the supply and cost of top-of-the-line FPGAs to early adopters.

To illustrate this impact consider a new process in which the defect density is 1/$cm^2$, which is a reasonable value early in the process lifecycle, and the die area is $6cm^2$, which roughly matches the size of the largest member of a high-end FPGA family such as Virtex 7. Using the Poisson Yield Model~\cite{yieldmodel}, the yield is only $0.25\%$ of die. If instead the chip is composed of four $1.5cm^2$ dice, the yield is $22\%$. This means that a 12 inch silicon wafer with $730cm^2$ of area would produce on average $0.3$ working $6cm^2$ dice, while the same wafer would produce on average $107$ working $1.5cm^2$ dice. Therefore, as a $6cm^2$ chip would be composed of four $1.5cm^2$ dice, the wafer would yield $26.75$ systems on average, as the "assembly yield" of placing these four die on an interposer is very high~\cite{xilinxTSV}. Hence the number of interposer-based FPGAs created from the same silicon wafer would be almost $100\times$ greater than a monolithic FPGA of the same size.

When the process matures and the defect density decreases this advantage drops significantly. Consider a mature process with defect density of $0.1/cm^2$. The yield for a $6cm^2$ die is $55\%$ and the yield for a $1.5cm^2$ die is $86\%$. Hence the number of single die FPGAs created from a $730cm^2$ silicon wafer would be $66.9$ and the number of interposer-based FPGAs created would be $104.6$. While the interposer-based FPGA still has a yield advantage it might not lead to a major cost advantage, particularly when the cost of the interposer and assembling the die to it are included. For the large, state-of-the-art FPGAs that are built early in a process cycle and heavily used for prototyping, however, there is clearly a compelling cost advantage to an interposer-based solution.

@@

The Xilinx 2.5D FPGAs from the Virtex-7 and Virtex Ultrascale families are currently the only commercially available silicon interposer-based FPGAs~\cite{xilinxWP}. As described in Section~\ref{archSection} we are studying the impact of several key interposer parameters on the performance of the multi-FPGA system, including (i) the percentage of the wiring normally present between rows of the FPGA which are cut when crossing between dice, and (ii) the extra delay (vs. a normal connection between adjacent rows) added when one must traverse the interposer.  To locate where Virtex-7 lies in this architecture space, we combined published information on the implementation of Xilinx's interposer-based FPGAs~\cite{xilinxTSV} with a detailed analysis of the~\textit{XC7V2000T} FPGA routing resources visible in the \textit{Vivado Device View}.

@@

The \textit{XC7V2000T} is composed of four identical dice arranged such that the vertical routing crosses between the dice. Each horizontal edge of each die has 280 groups of 48 length-12 wires crossing the interposer, which sums to a total of 13440 wires between dice. There are also 40 clock wires crossing the interposer. The average number of wires per vertical channel of this FPGA is 210 and there are approximately 280 vertical channels on the FPGA, resulting in approximately 58800 vertical wires crossing a horizontal cutline within a die. Hence the number of wires which cross the interposer is about 23\% of the total number of within-die vertical wires.

The 28nm dice are connected to the 65nm silicon interposer through microbumps with a 45$\mu$m pitch. Hence the area occupied by microbumps at one edge of one die is 13440 $\times$ (45$\mu$m)\textsuperscript{2} $=$ 27mm\textsuperscript{2}. If we assume each die is $7 \times 12 mm$, as presented by Chaware et al. in \cite{xilinxTSV}, the bumps have to be spread out near the edge and need to go as far as $2.25mm$ away from the edge of the die. This greater distance from the border increaes the length of the inter-die connections, and along with the presence of the micro-bumps and their capacitance, leads to an increased delay for these crossing wires vs. that of a typical on-die routing wire. Chaware et al. state that the latency to cross the interposer is approximately $1ns$. For comparison, a typical medium length 28 nm FPGA routing wire (e.g. spanning four logic blocks) has a delay of approximately 125 ps, while a longer wire (e.g. spanning 12 logic blocks) has a delay of approximately 250 ps.

Overall, these interposer-based FPGAs have increased delay and reduced connectivity between dice, with approximately $23\%$ of the usual number of vertical wires crossing between dice and approximately $1ns$ of increased delay to cross the interposer.

@@

Previous work by Pereira et al \cite{hahn2014cad} modeled the interposer indirectly: they did not introduce new rrnodes in the rrgraph to represent the interposer wires. Rather, they found the rrnodes that cross a cutline. Then they either removed all outgoing edges of these rrnodes (if the routing node was part of the \textit{\% wires cut}), or, they increased the switch delay of the rrnode fan-outs (if the wire was not cut) to account for the additional interposer delay.

@@

In~\cite{hahn2014cad}, Pereira et al tested several different \textit{cut\_cost} formulations, as well as different weighting for each. They found that the following \textit{cut\_cost} results in the best quality of results in terms of minimum routable channel width and critical path delay.

@@

This cost function performs better than several others as shown in \cite{hahn2014cad}, and we believe the key to the good performance of this cost functions is that \textit{cut\_cost} produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalizes bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

@@

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|r|r|}
\hline
 & Square Interposer & Square Dice \\ \hline \hline
bgm & 102 & 306 \\ \hline
stereovision0 & 86 & 256 \\ \hline
stereovision1 & 166 & 256 \\ \hline
LU8PEEng & 156 & 126 \\ \hline
mkDelayWorker32B & 96 & 194 \\ \hline
mcml & 196 & 496 \\ \hline
LU32PEEng & 256 & 264 \\ \hhline{|=|=|=|}
geomean & 141 & 252 \\ \hline
\end{tabular}
\caption{Minimum channel width for square interposer and square dice configurations.}
\label{table:squares}
\end{table}

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Conclusion}
\label{conclusionSection}
We have extended VPR to model and optimize for interposer-based multi-FPGA systems. While such systems are now a commercial reality, we do not know of any prior study of their key architectural parameters. We found that by modifying VPR's placement cost function we can improve routability (reduce minW), while simultaneously improving speed. Modifications to the VPR router lookahead function are crucial to efficiently routing interposer-based FPGAs, yielding a 31$\times$ speed-up. Incorporating partitioning into the CAD flow is very beneficial. Our best partitioning flow converts nets (hyperedges) to a star graph model, uses Metis to partition the resulting circuit graph, and forces VPR's packer to respect the Metis partitioning to FPGA dice. Taken together, these CAD changes reduce the signal wires that must cross the interposer by an average of 47\% while simultaneously improving circuit speed by 13\% for realistic interposer architectures.

We defined four key architecture parameters for interposer-based FPGAs, and used this extended VPR to analyze their impact on minimum channel width and critical path delay. We find that adding a moderate amount of switching flexibility at the FPGA - interposer boundary has negligible area cost but reduces the required FPGA minimum channel width by 13\%. We also find that the multi-die system has good routability so long as the interposer provides sufficient routing bandwidth, and this routing bandwidth can be much less than one would find in a monolithic FPGA. Specifically, if the interposer provides 40\% or more of the channel width required by the monolithic FPGA (an average of 45 wires per channel across our benchmark suite) the routability of the interposer-based FPGA is equivalent to that of a monolithic one. As the interposer signal bandwidth becomes more constrained, there is some impact to routability, but even an interposer providing only 20\% of the monolithic FPGA routing channel bandwidth (22 wires in our experiments) only needs a 10\% increase in the within-FPGA-die routing channel width to compensate.

The \textit{critical path delay} is not strongly influenced by the \textit{\% wires cut} but is strongly influenced by the interposer delay and the number of cuts. Our results show that the critical path often crosses the interposer cutline more than once, as the \textit{critical path delay} is increased by more than the interposer \textit{delay increase}; on average we find that a 1 ns interposer delay causes an 11\% delay increase in a two-die system vs. a monolithic FPGA . Increasing the number of dice in an interposer-based FPGA does not greatly impact the \textit{minimum channel width}, but does lead to a larger \textit{critical path delay}. We also found that the aspect ratio of the interposer-based FPGA is an important architectural parameter; for the best routability one should use a square interposer and rectangular FPGA dice.

Overall we find the results for interposer-based FPGAs to be very positive. While an interposer can only provide a fraction of the signal count between dice that one would have across cutline within a die, the system still has excellent routability and behaves similarly to a monolithic FPGA so long as the interposer signal count is not pushed to very low values; below 20\% of the within-die routing in our experiments. We have also found that modifications to the FPGA CAD flow and routing architecture to exploit interposer-based FPGAs have been very productive. As this is a new area we believe more routability and timing enhancements are likely with further research, and this would further enhance the utility of interposer-based FPGAs.

% use section* for acknowledgement
\section*{Acknowledgment}
This work was supported by a Ci\^{e}ncia sem Fronteiras scholarship from CNPq - Brazil and the NSERC/Altera Industrial Research Chair in Programmable Silicon. Some of our computations were performed on the GPC supercomputer at the SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for Innovation under the auspices of Compute Canada; the Government of Ontario; Ontario Research Fund - Research Excellence; and the University of Toronto.

>>>>

\author{
		% 1st author
		Ehsan~Nasiri,
		% 2nd author
		Javeed~Shaikh,
		% 3rd author
		Andre~Hahn~Pereira,
		% 4th author
        and~Vaughn~Betz,~\IEEEmembership{Member,~IEEE}%
\thanks{E. Nasiri is with Altera, Toronto, ON, Canada (email: enasiri@altera.com)}% 
\thanks{J. Shaikh is with Google, Mountain View, CA, USA (email: jvds@google.com)}%
\thanks{A. H. Pereira is with the Computer and Digital Systems Engineering Dept., University of S\~{a}o Paulo, SP, Brazil (email: andre.hahn@usp.br)}%
\thanks{V. Betz is with the Dept. of Electrical and Computer Engineering, University of Toronto, ON, Canada (email: vaughn@eecg.utoronto.ca)}%
\thanks{Manuscript received January 6, 2015.}}

@@

% As a general rule, do not put math, special symbols or citations in the abstract or keywords.
\begin{abstract}
Multiple FPGA dice can be connected through a silicon interposer to form a larger system than one monolithic die can accommodate. Such systems are now commercially available but many open questions remain concerning their key architecture parameters and efficiency, as the signal count between dice is reduced and the signal delay between dice is increased compared to a monolithic FPGA.
We create a new version of VPR to target interposer-based FPGAs and investigate modifications to placement and routing and the incorporation of partitioning into the flow to improve results. Our best CAD flow reduces the routing demand for interposer FPGAs with realistic amounts of connectivity between dice by 47\% and improves the circuit speed by 13\% on average. Architecture modifications to add routing flexibility when crossing the interposer are very beneficial and improve routability by a further 11\%. With these CAD and architecture enhancements, we find that if an interposer supplies (between dice) 40\% of the routing capacity that the normal (within-die) FPGA routing channels supply there is negligible impact on circuit routability. Smaller interposer routing capacities do impact routability however: minimum channel width increases by 10\% and 70\% when an interposer supplies only 20\% and 10\% of the within-die routing, respectively. The interposer also impacts delay, increasing circuit delay by 11\% on average for a 1 ns interposer signal delay and a two-die system. 
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Algorithms, FPGA, Silicon interposer, 2.5D ICs. 
\end{IEEEkeywords}

\section{Introduction}
\label{introSection} 
%   - Describe multi-chip
Interposer-based FPGA systems are composed of multiple FPGA dice connected through a silicon interposer. The interposer uses an older fabrication technology than the FPGA dice, and links between dice include both a micro-bump on each FPGA die and a metal wire on the interposer~\cite{xilinxTSVperformance}. This results in reduced connectivity and increased delay between dice compared to the routing bandwidth and delay across a cutline within a single die. Such interposer systems are sometimes called 2.5D devices, since they make use of vertical stacking of dice on an interposer to enable higher integration levels~\cite{3dhandbook}. In this work we present an architecture study of silicon interposer-based FPGAs that analyzes how the reduced connectivity and increased delay impact their performance.

Silicon interposers enable the creation of large FPGAs from smaller dice, which can both improve yield and fabricate FPGAs that are too large to be manufacturable on a single die. Making large FPGAs from multiple smaller dice is particularly interesting at the beginning of a new manufacturing process, when defect densities are high. In this case, good-die yield drops dramatically as the die size increases, drastically impacting the availability of large FPGAs early in the process lifetime.

@@

%   - Contributions
The main contributions of our work are as follows:
\begin{itemize}
	\item VPR CAD tool modifications to model and optimize for 2.5D silicon interposer-based FPGAs.
	\item An investigation of the efficacy of incorporating partitioning into various stages of the CAD flow.
	\item A study of how multi-FPGA system routability is impacted by the amount of between-die connectivity and by the switching flexibility at the FPGA die - interposer boundary.
	\item An analysis of the interposer's impact on timing.
\end{itemize}

The paper is organized as follows. Section \ref{siliconSection} gives more information about silicon interposer technology. Section \ref{archSection} describes the changes made to the FPGA architecture description and VPR to target 2.5D FPGAs. Section \ref{cadSection} describes optimizations to the VPR placement and routing algorithms for interposer FPGAs and Section~\ref{sec:CADflow} describes how partitioning can further enhance the CAD flow. Section \ref{resultsSection} presents architectural results for 2.5D FPGAs.

An earlier and less detailed version of this work appeared in \cite{interposer2014}. We greatly extend this earlier work by enhancing the CAD tools, incorporating partitioning into the flows and investigating a broader range of architecture questions.

@@

Interposer-based FPGAs allow systems larger than a single die, making a "More than Moore" improvement on the size and number of logic cells possible, and with chips combined with far more connectivity and less delay than if they were in separate packages connected through a printed circuit board.

The improvement in the capacity of 2.5D FPGAs over conventional ones is very significant. Xilinx's largest 28 nm interposer-based FPGA contains 4 FPGA dice and 1.954 million logic cells~\cite{xilinx7series}, while the largest non-interposer Virtex-7 die has 979k logic cells and Altera's largest 28 nm FPGA has 952k logic elements~\cite{stratixV}. Even though all these FPGAs use a 28 nm process, silicon interposer technology allows the creation of FPGAs with twice the resources possible on even an extremely large single die. Xilinx's recently announced Virtex UltraScale line of 20 nm FPGAs makes even more aggressive use of interposers; the largest monolithic FPGA in this family contains 941k logic cells, while the four largest capacity FPGAs are all built with silicon interposers and span a range from 1.25 to 4.43 million logic cells~\cite{virtexultrascale}.

Another major advantage of interposer-based FPGAs comes at the beginning of a new manufacturing process, when the defect density is high~\cite{xilinxTSVperformance}. To illustrate this impact consider a new process in which the defect density is 1/$cm^2$ (a reasonable value early in the process lifecycle) and the die area is $6cm^2$, which roughly matches the size of the largest member of a high-end FPGA family such as Virtex 7. Using the Poisson Yield Model~\cite{yieldmodel}, the yield is only $0.25\%$ of dice. If instead the chip is composed of four $1.5cm^2$ dice, the yield is $22\%$. This means that a 12 inch silicon wafer with $730cm^2$ of area would produce on average $0.3$ working $6cm^2$ dice, while the same wafer would produce on average $107$ working $1.5cm^2$ dice. Therefore, as a $6cm^2$ chip would be composed of four $1.5cm^2$ dice, the wafer would yield $26.75$ systems on average, as the "assembly yield" of placing these four die on an interposer is very high~\cite{xilinxTSV}. Hence the number of interposer-based FPGAs created from the same silicon wafer would be almost $100\times$ greater than a monolithic FPGA of the same size, greatly impacting not only cost but also the availability of such a large FPGA.

When the process matures and the defect density decreases this yield advantage drops. Consider a mature process with defect density of $0.1/cm^2$. The yield for a $6cm^2$ die is $55\%$ and the yield for a $1.5cm^2$ die is $86\%$. Hence the number of single die FPGAs created from a $730cm^2$ silicon wafer would be $66.9$ and the number of interposer-based FPGAs created would be $104.6$. While the interposer-based FPGA still has a yield advantage it might not lead to a major cost advantage when the cost of the interposer and assembling the die to it are included. For the large, state-of-the-art FPGAs that are built early in a process cycle and heavily used for prototyping, however, there is clearly a compelling cost advantage to an interposer-based solution.

@@

The Xilinx 2.5D FPGAs from the Virtex-7 and Virtex Ultrascale families are currently the only commercially available silicon interposer-based FPGAs~\cite{xilinxWP}. As described in Section~\ref{archSection} we are studying the impact of several key interposer parameters on the performance of the multi-die system, including (i) the percentage of the wiring normally present between rows of the FPGA which is cut when crossing between dice, and (ii) the extra delay (vs. a normal connection between adjacent rows) added when one must traverse the interposer.  To locate where Virtex-7 lies in this architecture space, we combined published information on the implementation of Xilinx's interposer-based FPGAs~\cite{xilinxTSV} with a detailed analysis of the~\textit{XC7V2000T} FPGA routing resources visible in the \textit{Vivado Device View}.

@@

The \textit{XC7V2000T} is composed of four identical dice arranged such that the vertical routing crosses between the dice. Each horizontal edge of each die has 280 groups of 48 length-12 wires crossing the interposer, which sums to a total of 13440 wires between dice. There are also 40 clock wires crossing the interposer. The average vertical routing channel in this FPGA contains 210 wires and there are approximately 280 vertical channels, resulting in approximately 58800 vertical wires crossing a horizontal cutline within a die. Hence the number of wires which cross the interposer is about 23\% of the total number of within-die vertical wires.

The 28 nm dice are connected to the 65 nm silicon interposer through microbumps with a 45$\mu$m pitch. Hence the area occupied by microbumps at one edge of one die is 13440 $\times$ (45$\mu$m)\textsuperscript{2} $=$ 27mm\textsuperscript{2}. If we assume each die is $7 \times 12 mm$, as presented by Chaware et al. in \cite{xilinxTSV}, the bumps have to be spread out near the edge and need to go as far as $2.25mm$ away from the edge of the die. This greater distance from the border increase the length of the inter-die connections, and along with the presence of the micro-bumps and their capacitance, leads to an increased delay for these crossing wires vs. that of a typical on-die routing wire. Chaware et al. state that the latency to cross the interposer is approximately $1ns$. For comparison, a typical medium length 28 nm FPGA routing wire (e.g. spanning four logic blocks) has a delay of approximately 125 ps, while a longer wire (e.g. spanning 12 logic blocks) has a delay of approximately 250 ps.

Overall, these interposer FPGAs have increased delay and reduced connectivity between dice, with approximately $23\%$ of the usual number of vertical wires between dice and approximately $1ns$ of increased delay to cross the interposer.

@@

Previous work by Pereira et al \cite{interposer2014} modeled the interposer indirectly: they did not introduce new rrnodes in the rrgraph to represent the interposer wires. Rather, they found the rrnodes that cross a cutline. Then they either removed all outgoing edges of these rrnodes (if the routing node was part of the \textit{\% wires cut}), or, they increased the switch delay of the rrnode fan-outs (if the wire was not cut) to account for the additional interposer delay.

@@

In~\cite{interposer2014}, Pereira et al tested several different \textit{cut\_cost} formulations, as well as different weighting for each. They found that the following \textit{cut\_cost} results in the best quality of results in terms of minimum routable channel width and critical path delay.

@@

This cost function performs better than several others as shown in \cite{interposer2014}, and we believe the key to the good performance of this cost functions is that \textit{cut\_cost} produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalizes bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

@@

\begin{table}[!htbp]
\centering
\caption{Square interposer and square dice minimum channel widths}
\begin{tabular}{|l|r|r|}
\hline
 & Square Interposer & Square Dice \\ \hline \hline
bgm & 102 & 306 \\ \hline
stereovision0 & 86 & 256 \\ \hline
stereovision1 & 166 & 256 \\ \hline
LU8PEEng & 156 & 126 \\ \hline
mkDelayWorker32B & 96 & 194 \\ \hline
mcml & 196 & 496 \\ \hline
LU32PEEng & 256 & 264 \\ \hhline{|=|=|=|}
geomean & 141 & 252 \\ \hline
\end{tabular}
\label{table:squares}
\end{table}

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Conclusion}
\label{conclusionSection}
We have extended VPR to target interposer-based multi-FPGA systems. We found that by modifying VPR's placement cost function we can improve routability, while simultaneously improving speed. Modifications to the VPR router lookahead function are crucial to keep compile time reasonable and yield a 31$\times$ speed-up. Incorporating partitioning into the CAD flow is very beneficial. Our best partitioning flow converts nets (hyperedges) to a star graph model, uses Metis to partition the resulting circuit graph, and forces VPR's packer to respect the Metis partitioning to FPGA dice. Taken together, these CAD changes reduce the signal wires that must cross the interposer by an average of 47\% while simultaneously improving circuit speed by 13\% for realistic interposer architectures.

We defined four key architecture parameters for interposer-based FPGAs, and used this extended VPR to analyze their impact on minimum channel width and critical path delay. We find that adding a moderate amount of switching flexibility at the FPGA - interposer boundary has negligible area cost but reduces the required minimum channel width by 13\%. We also find that the multi-die system has good routability so long as the interposer provides sufficient routing bandwidth, and this routing bandwidth can be much less than one would find in a monolithic FPGA. Specifically, if the interposer provides 40\% or more of the channel width required by the monolithic FPGA (an average of 45 wires per channel across our benchmark suite) the routability of the interposer-based FPGA is equivalent to that of a monolithic one. As the interposer signal bandwidth becomes more constrained, routability is impacted but even an interposer providing only 20\% of the monolithic FPGA routing channel bandwidth (22 wires in our experiments) only needs a 10\% increase in the within-FPGA-die routing channel width to compensate.

The \textit{critical path delay} is not strongly influenced by the \textit{\% wires cut} but is strongly influenced by the interposer delay and the number of cuts. On average we find that a 1 ns interposer delay causes an 11\% delay increase in a two-die system vs. a monolithic FPGA. Increasing the number of dice in an interposer-based FPGA only moderately impacts the \textit{minimum channel width}, but does lead to a larger \textit{critical path delay}. The aspect ratio of the interposer-based FPGA is an important architectural parameter; a square interposer containing rectangular FPGA dice has the best routability.

Overall we find the results for interposer-based FPGAs to be very positive. While an interposer can only provide a fraction of the signal count between dice that one would have across cutline within a die, the system still has excellent routability so long as the interposer signal count is not pushed to very low values; below 20\% of the within-die routing in our experiments. We have also found that modifications to the CAD flow and routing architecture to exploit interposer-based FPGAs have been very productive. As this is a new area we believe more routability and timing enhancements are likely with further research; for example, a timing-driven partitioner may reduce the delay impact of the interposer.

% use section* for acknowledgement
\section*{Acknowledgment}
This work was supported by Brazil's Ci\^{e}ncia sem Fronteiras scholarship and the NSERC/Altera Industrial Research Chair in Programmable Silicon. Some computations were performed on the GPC supercomputer at the SciNet HPC Consortium.
